Exploring the Role of Feedback in Machine Learning Jupyter Notebooks

### 2024-05-07 report/emse24.tex
feat(report): emse24 updates

\begin{abstract}
Insert your abstract here. Include keywords, PACS and mathematical
subject classification numbers as needed.
\keywords{First keyword \and Second keyword \and More}
% \PACS{PACS code1 \and PACS code2 \and more}
% \subclass{MSC code1 \and MSC code2 \and more}
\end{abstract}

@@

\section{Methodology}
% For one-column wide figures use

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{esem24-method.pdf}
  \label{fig:data-collection}
  \caption{Overview of data collection methodology used in this study}
\end{figure}

@@

\subsection{Implicit feedback using \texttt{print} statements and output of code cells}

\subsubsection{Data distribution check (O2, O4, O9, O14, O20, O25, O48)}
% a mix of visualisations and pandas dataframes that show the distribution of a column in the data
% we primarily observe these checks being used to determine the relationship of a feature with the target feature
% typically in the EDA/data exploration/understanding phase
% several examples where the distribution of a newly created feature is checked (binning a continuous feature into categorical one)
% one uses descriptive statistics to verify that data normalisation worked (again, a check after data transformation step)

@@

\bibliographystyle{ACM-Reference-Format} \bibliography{bibliography}
%\begin{acknowledgements}
%If you'd like to thank anyone, place your comments here
%and remove the percent signs.
%\end{acknowledgements}

@@

% BibTeX users please use one of
\bibliographystyle{spbasic}      % basic style, author-year citations
%\bibliographystyle{spmpsci}      % mathematics and physical sciences
%\bibliographystyle{spphys}       % APS-like style for physics
\bibliography{bibliography}   % name your BibTeX data base

>>>>

\begin{abstract}

  % TODO update the numbers here
  % TODO: update
Testing ML systems is a highly interactive process which demands a human-in-the-loop approach. In addition to writing tests for the code base, practitioners are required to analyse and interpret several visualisations using their domain expertise to validate if an ML system satisfies the required set of functional and non-functional properties. Visualisations are frequently used to qualitatively assess various parts of an ML pipeline. However, implicit knowledge gained from visualisations must be translated to explicit analytical tests that fail when there is change in any component of the ML pipeline. We conduct an empirical analysis of Jupyter notebooks to catalogue the state-of-the-art mappings between ML visualisations and assertions. We mine Github to collect 54K Jupyter Notebooks that contain assertions written in Python. We develop a novel methodology to identify 1764 notebooks which contain an assertion below a visualisation. We manually analyse the 1.7K notebooks and identify 269 visualisation-assertion pairs that are semantically related to one another. We further investigate the 269 visualisation-assertion pairs and identify three frequently occurring testing patterns. We perform an in-depth analysis of 34 visualisation-assertion pairs and find that the assertions often fail to capture all the information present in their visual counter-part. Empirical evidence obtained in this study indicates that current software testing methods fail to address the unique challenges of ML. And emphasises the need for automated tools that bridge the gap between visual assessments and analytical assertions.

\keywords{ML Testing \and SE4AI \and Visualisations \and Assertions \and Computational Notebooks}
\end{abstract}

@@

% amershi2019software: start with ML development life cycle and how everything is done in a very iterative way; things are dependant on one another; etc.

% martinez-plumed2021crisp-dm: not only does ML include the crip-dm framework, but this same form of iteration is seen in the model development and experimentation phase;

% this is why we see the use of computational notebooks being widely adopted by the ML community

% writing ML code is more iterative and requires a constant feedback; we identify this feedback can be provided in two ways: implicitly from the output of a cell and explicitly from a failed assertion

Code, data and ML model in ML-enalbed systems are highly tangled with each other. For instance, the decision to use a particular type of ML model, requires a deep understanding of the data which will be used to train the model. Data quantity and quality directly influence the complexity of the ML model.

Conversely, the choice of the ML model also influences the data pre-processing and transformation steps that are required to get the data ready for training. For instance, using a classic statistical model requires several data transformation steps such as scaling and normalization, and converting categorical features to numerical ones using one-hot encoding.

Visualisations are used throughout the ml development lifecycle to test various properties of the ml system. in the early stages of the ml development lifecycle, visualisations are extensively used to make sense of the data and verify its statistical properties. during the model development phase, visualisations are used to summarise metrics, contrast different learning algorithms and iteratively fine-tune ml models. once the model is deployed in production, visualisations are used to continually monitor their performance and trigger a new training cycle once their performance drops below a certain threshold~\cite{yuan2021survey,hohman2019visual,amershi2015modeltracker,wexler2020if}.

building ml systems is highly iterative and experimental. information gained from ml visualisations is used to make design and implementation decisions for the following steps of the ml pipeline. for instance, we may visualise the distribution of our training data and find that it is normally distributed. based on this information, we may opt for a linear regression model which assumes normality in the underlying data. however, such expectations regarding the data may be violated once the ml system is deployed in production, where the data constantly changes as a reflect of the real world~\cite{amershi2019software,sambasivan2021everyone,breck2019data,baylor2017tfx}.

implicit expectations obtained from visualisations must therefore be
translated to analytical tests that fail once our expectations
regarding the ml system are no longer satisfied.

in contrast to prior work, we approach ml testing from a new
perspective. we conduct an empirical analysis of computational
notebooks obtained from github, to understand the process of testing
ml systems in practice. in particular, we focus on the combination of
a qualitative form of testing (using visualisations) and a
quantitative form of testing (using analytical assertions). the
research questions along with the contributions of this paper are as
follows.

\begin{description}
  \item[rq1.] \textbf{how frequently are analytical tests formulated
    from visualisations created to test ml systems?}

    we mine 54k jupyter notebooks from github that contain an assertion
    written in python. we develop an automated technique to identify 1764
    notebooks with an assertion below a visualisation. we manually analyse
    the 1.7k notebooks and identify 269 visualisation-assertion (va) pairs
    that are semantically related to each other. we plan to release this
    catalogue of va pairs publicly.

\item[rq2.] \textbf{what patterns are frequently observed in va pairs
  used to test ml systems?}

  we analyse the 269 va pairs from rq1, and observe three testing
  patterns that frequently occur in the va pairs.

\item[rq3.] \textbf{do the assertions capture all the information
  obtained from the corresponding visualisation?}

  we further perform an in-depth analyse of 34 va pairs (and their
  corresponding notebooks). we observe that assertions often cannot
  capture all the information obtained from the visualisation.

\end{description}

our observations indicate that formulating analytical assertions from
visualisations is an emerging testing technique. however, the
assertions found in this study often do not reflect the information
obtained from their visual counter-part. this study finds the need for
automated tools that can reduce the manual effort necessary to
validate visualisations in ml systems. and help ml practitioners to
formulate better analytical assertions from visualisations.

the replication package for this study is made public on
figshare\footnote{the replication package can be found at this url:
https://figshare.com/s/9347228011999cba29f5}.

\section{Preliminaries}\label{sec:prelim}

This section summarises prior scientific work that has been done in ML
Testing, computational notebooks and visual analytics. The section
concludes with an overview of technical knowledge required for this
paper.

\subsection{ML Testing}\label{sec:ml-testing}

Testing ML enabled software systems poses more challenges compared to
their traditional counter part. While traditional software systems
mature through change in their codebase, ML enabled systems
additionally experience change in the training data and the machine
learning
model~\cite{sculley2015hidden,amershi2019software,sambasivan2021everyone}.
With ML being adopted into safety-critical domains that can affect
human lives, ensuring that ML systems are correct, robust to data
perturbations and not biased by race, colour or gender is of paramount
importance. Existing scientific literature on ML testing broadly
focuses on two aspects. First, on functional properties such as
correctness and robustness of the model towards unseen data. And
second, on non-functional properties such as fairness,
interpretability and
privacy~\cite{zhang2020machine,mehrabi2021survey,chen2022fairness}.

To test the correctness of ML models, several improvements over
existing test adequacy metrics have been proposed. Tools such as
\textit{DeepExplore} and \textit{DeepGauge} propose new test adequacy
metrics such as neuron coverage adapted for ML enabled software
systems~\cite{pei2017deepexplore, ma2018deepgauge,
  gerasimou2020importance}. Formal verification methods have also been
proposed that try to provide formal guarantee of robustness against
adversarial examples. Such methods however are only feasible for
statistical ML models and become computationally expensive for more
complex models such as deep neural networks~\cite{zhu2021deepmemory,
  baluta2021scalable}. Several works have been conducted on generating
and detecting adversarial inputs for ML models. Data augmentation
techniques based on fuzzing, search based software testing and
mutation testing have been proposed to generate adversarial examples
that can be used during model training to improve its
robustness~\cite{braiek2019deepevolution, gao2020fuzz, wang2021robot,
  zhang2020white}. It is however not possible to include all
variations of adversarial examples into the training data. Thus,
methods have been proposed to detect adversarial inputs during
runtime~\cite{xiao2021self, wang2020dissector, wang2019adversarial,
  berend2020cats}.

In contrast to existing scientific contributions, we conduct a
large-scale empirical analysis of Jupyter Notebooks to understand the
process of testing ML systems. Specifically, we focus on how
visualisations are used to test specific properties of ML systems. We
further investigate how frequently analytical tests are formulated
based on the information gained from the visualisations.

\subsection{Computational Notebooks and Software Engineering}\label{sec:notebooks}

Computational notebooks have been ubiquitously adopted by the machine
learning community for developing ML enabled systems. Although
originally intended to promote reproducible software, computational
notebooks are far from being reproducible due to lack of software
engineering best practices such as separation of concern, testing and
versioning~\cite{pimentel2019large,wang2020better,chattopadhyay2020wrong}.

\emph{Psallidas et al} provide an overview of the evolving landscape
of computational notebooks by analysing six million Python notebooks,
two million enterprise Data Science (DS) pipelines, source code and
metadata from over 900 releases of 12 important DS libraries. Their
findings can be used by system builders for improving DS tools and
also by DS practitioners to understand the current trends in
technologies they should focus on. \emph{Pimentel et al} mined 1.4
million notebooks from Github to conduct an empirical study on the
coding practices in computational notebooks. Based on their analysis,
the authors propose guidelines on improving reproducibility of
computational notebooks. To enable future research on computational
notebooks, \emph{Quaranta et al} mine Kaggle and present
\textit{KGTorrent}, a public dataset consisting of approximately 2.5
million Jupyter notebooks written in Python. The dataset also contains
a relational database dump of metadata regarding publicly available
notebooks on Kaggle.

Studies with human subjects have been conducted to gain a deeper
understanding of the challenges faced by ML practitioners when
developing ML models inside notebooks. The results indicate that ML
practitioners work in a highly iterative fashion, often experimenting
with multiple strategies to analyse the data or produce meaningful
visualisations~\cite{kandel2012enterprise, kery2018story,
  liu2019understanding, chattopadhyay2020wrong}. Studies have also
been conducted to understand how practitioners generate, evaluate and
manage alternative hypothesis, visual designs, methods, tools,
algorithms and data sources to arrive at the final
implementation~\cite{liu2019understanding,kandel2012enterprise}. The
findings from these studies can be used as guidelines for improving
existing notebook technologies or designing new ones.

To manage and prune multiple versions of code that accumulate over
time when developing in notebooks, \emph{Head et al} propose a code
gathering tool that allows practitioners to review and only keep the
relevant version of code. Other tools such as \textit{WrangleDoc} and
\textit{VizSmith} have also been proposed to aid ML practitioners when
working within computational notebooks~\cite{yang2021subtle,
  bavishi2021vizsmith}.

Jupyter notebooks have been widely adopted by the DS and ML
communities to develop ML
pipelines~\cite{wang2020assessing,pimentel2019large,quaranta2021kgtorrent}.
Computational notebooks provide a rich source of data in the form of
natural text, code and visualisations. Computational notebooks also
allow ML practitioners to present the knowledge gained from the
analysis in a narrative that can be shared with
others~\cite{rule2018exploration}. This study leverages the
computational narrative present within notebooks to identify
visualisations and assertions that are semantically linked to each
other.

\subsection{Visualisations and Machine Learning}\label{sec:visualisations}

As ML augment software systems in safety-critical domains, emphasis
has been put into explainability of ML models. ML models and the
underlying data is complex and multi-dimensional. To combat the
``curse of dimensionality'', visual analytics has been widely adopted
by the ML community to understand the data and the internal workings
of ML models~\cite{yuan2021survey,hohman2019visual,wexler2020if}.

Prior studies have been conducted to understand how visual analytics
techniques are currently being used in ML. \emph{Yuan et al} conduct a
systematic review of 259 papers and propose a taxonomy of visual
analytics techniques for ML. \emph{Hohman et al} conduct a survey of
visual analytics techniques for Deep Learning Models. The findings of
the study indicate that visual analytics has been widely adopted in ML
for model explanation, interpretation, debugging and improvement.

Several tools have been proposed by the visual analytics research
community to aid practitioners in understanding how their ML models
operate. \emph{Wexler et al} propose \textit{What-If}, a visual
analytics tool to explore alternative hypotheses, generate
counterfactual reasoning, investigate decision boundaries of the model
and how change in data affects the model predictions. To reduce the
cognitive load of ML practitioners during model building phase,
\emph{Amershi et al} propose \textit{ModelTracker}. Given a trained
model and a sample dataset, ModelTracker presents all the information
in traditional summary statistics along with the performance of the
model.

\textit{ESCAPE}, \textit{GAM Coach}, \textit{Angler} and
\textit{Drava} are a few other tools that have been proposed to handle
specialised use-cases such as identifying systematic errors in ML
models, generating counterfactual explanations for Generalised
Additive Models, prioritising machine translation model improvements
and relating human concepts with semantic dimensions extracted by ML
models during disentangled representation learning\cite{ahn2023escape,
  wang2023gam, robertson2023angler, wang2023drava}.

In contrast to prior work which propose dedicated visual analytics
tools, this study focuses on visualisations within computational
notebooks. We focus on visualisations produced by AI practitioners
using Python libraries (such as Matplotlib or Seaborn). Additional
context provided by surrounding code and markdown cells are used to
understand the motivation for creating the visualisation and the
information gained from it.

\section{Study Design}

% TODO: introduce the motivation of the study along with the RQs here

% role of feedback in writing ML code? (any recommendation for papers I can refer to here?)
% implicit vs. explicit feedback (how do we establish this concept?)

\subsection{Data Collection}\label{sec:data-collect}

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{data-collection.pdf}
  \label{fig:data-collection}
  \caption{Overview of data collection methodology used in this study}
\end{figure}

Figure~\ref{fig:data-collection} summarises the data collection procedure used in this study. We collected Jupyter notebooks written in Python, from Kaggle and Github.

We mined public repositories from Github to collect $49923$ Jupyter notebooks. We use the Github advanced search syntax~\footnote{https://docs.github.com/en/search-github/searching-on-github/searching-code} to isolate Jupyter Notebook that contain the keyword ``assert'' in them. The final search query is as follows: \texttt{"assert" language:"Jupyter Notebook"}.\footnote{Collected on June 22, 2023}

For Kaggle, we used a pre-existing dataset KGTorrent~\cite{quaranta2021kgtorrent} since Kaggle does not support advanced code-based search like Github. To the best of our knowledge, KGTorrent this is the largest dataset of Python Jupyter notebooks obtained from Kaggle consisting of $248763$ notebooks.

Therefore, we start with 283GB of data comprising of $298686$ Python Jupyter notebooks.

We check the validity of the underlying JSON structure in each notebook using the nbformat tool ~\footnote{https://nbformat.readthedocs.io/en/latest/}. Since the focus of this study is on the analysis of Python code, we only include notebooks that contain at least one code cell. Finally, to focus on Python code written specifically for ML projects, we only include notebooks that import popular ML libraries~\footnote{These libraries are derived from white literature---https://www.coursera.org/articles/python-machine-learning-library, https://www.geeksforgeeks.org/best-python-libraries-for-machine-learning, https://www.datacamp.com/blog/top-python-libraries-for-data-science, https://www.kdnuggets.com/2020/11/top-python-libraries-data-science-data-visualization-machine-learning.html, https://lp.jetbrains.com/python-developers-survey-2022}.

For explicit feedback using assertions, we programmatically analyse the JSON structure of the notebooks to isolate code cells. We collect Python assert statements from the code cells by constructing and subsequently parsing the Abstract Syntax Tree (AST) of the source code present in each cell.

Similarly, for implicit feedback, we first isolate code cells that produce an output. The output of a cell may either be from a ``print'' statement or the last statement of the cell. We collect both of them using the AST of the cell.

Finally, we remove duplicate data points (asserts, prints and last statements) resulting in a final data set of $22548$ assertions, $25319$ print statements and $28522$ last statements.

\subsection{Case Studies}

% TODO: might need a visual aid here

We allocated a fixed time resource of 100 hours to conduct the case study analysis of all candidates (assertions and cell outputs). To surface interesting candidates for the case studies, we apply NLP techniques as described below.

% TODO: point out the poor quality of notebooks out there (we have citations for this); and the resulting noise in the assertions/outputs. Thats why we need to apply these tecniques to surface unique and interesting candidates!

The assertions and cell outputs are first tokenised---special characters and alpha-numeric words shorter than two characters are removed. Two stop words namely ``assert'' and ``print'' are removed since they appear in all assertions and print statements respectively. The term frequency (TF) for all tokens is calculated and then normalised using their inter-document frequency (IDF) such that tokens thay appear less frequently are assigned a higher value.

We apply stratified random sampling to identify the candidates for the case study analysis. The sub-groups are created by adding TF-IDF of the tokens in each candidate to produce an aggregate value. The candidates are then divided into quartiles based on the aggregate value. A candidate is randomly drawn from each bin and analysed as an in-depth case study. The analysis is stopped when the time resource is exhausted.

During each case study, we analyse the code of the candidate to understand its purpose. Additionally, we analyse the entire code cell, the previous cell, next cell and the notebook's purpose to bring in rich context.

@@

\subsection{Implicit feedback from the output of code cells}

\subsubsection{Data distribution check ($N = 7$)}
% (O2, O4, O9, O14, O20, O25, O48)

% we need to know the distribution of the data to make informed decisions regarding the data transformation steps required (for instance, do we need to scale and normalise? do we have outliers?); visualisations are a cheap and quick means of doing this

% distribution check is also required since models such as linear regression, perform well when the data is normally distributed; if our data is not normal then we need to perform additional transformation steps to make the data ready for training

% a mix of visualisations and pandas dataframes that show the distribution of a column in the data
% we primarily observe these checks being used to determine the relationship of a feature with the target feature
% typically in the EDA/data exploration/understanding phase
% several examples where the distribution of a newly created feature is checked (binning a continuous feature into categorical one)
% one uses descriptive statistics to verify that data normalisation worked (again, a check after data transformation step)

@@

%\begin{acknowledgements}
%If you'd like to thank anyone, place your comments here
%and remove the percent signs.
%\end{acknowledgements}

@@

% BibTeX users please use one of
% \bibliographystyle{spbasic}      % basic style, author-year citations
\bibliographystyle{spmpsci}      % mathematics and physical sciences
%\bibliographystyle{spphys}       % APS-like style for physics
\bibliography{bibliography}   % name your BibTeX data base
